The locally Gaussian partial correlation

### 2020-09-02 lgpc_JBES.tex
Implementing comments from Dag and one referee

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The locally Gaussian partial correlation},
  pdfauthor={Håkon Otneim and Dag Tjøstheim},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}

@@

\definecolor{Gray}{gray}{0.9}

@@

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle
\begin{abstract}
It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally, we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
\end{abstract}

@@

Let \(\X = (X_1, \ldots, X_p)\) be a random vector. Denote by \((\X_1,\X_2,\X_3)\) a partition of \(\X\) into vectors of dimension \(d_1\), \(d_2\) and \(d_3\) respectively, such that \(\Xone = (\X_1, \X_2) = (X_1, \ldots,X_{d_1 + d_2})\) consists of the first \(d_1+d_2\) components in \(\X\) and \(\Xtwo = \X_3 = (X_{d_1+d_2+1}, \ldots, X_{d_1+d_2+d_3})\) contains the remaining \(d_3\) variables, where \(p = d_1+d_2+d_3\). We assume that the mean vector \(\fmu\) and covariance matrix \(\fSigma\) of \(\X\) exist, and partition them correspondingly, writing
\begin{equation}
\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix} \,\,\,\, \textrm{ and } \,\,\,\, \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33}\end{pmatrix},
\label{eq:matrixpartition}
\end{equation}
where \(\fSigma_{ij}\) is the covariance matrices of \((\X_i, \X_j)\), \(i,j = 1,2,3\). There are two main concepts of correlation when \(\Xtwo = \X_3\) is given, the \emph{partial} and the \emph{conditional} correlation, and they coincide in several joint distributions, among them the Gaussian. See e.g.~Baba, Shibata, and Sibuya (\protect\hyperlink{ref-baba2004partial}{2004}) for details. We will use the partial correlation as a starting point when defining the LGPC. The partial variance-covariance matrix of \(\Xone = (\X_1, \X_2)\) given \(\Xtwo = \X_3\) is
\begin{equation}
\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{-1}\right)^{22}\fSigma^{21},
\label{eq:matrixpartial}
\end{equation}
where
\[\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22} \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix}, \,\,\fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix} \, \textrm{ and } \, \fSigma^{22} = \fSigma_{33},\]
and \(\fSigma_{12|3}\) is the covariance matrix in the conditional (Gaussian) distribution of \(\Xone\) given \(\X_3\) if \(\X\) is jointly normal. The partial correlation matrix between \(\X_1\) and \(\X_2\) given \(\X_3\) is naturally defined as

@@

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
\label{eq:trans}
\end{equation}
The transformation enables us to simplify the local Gaussian representation \eqref{eq:localgaussian} by writing the density \(f_{\Z}\) of \(\Z\) as
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
\label{eq:lgdeapprox}
\end{equation}\\
where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a simplified representation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.

@@

\[f_{z_1|\z_3}(z_1) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
where

@@

\begin{equation}
\hR(\z) = \arg\max_{\R} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \R) - \int K_{\hh}(\y - \z)\psi(\y, \R)\,\textrm{d}\y
\label{eq:locallik3}
\end{equation}
in each point \(\z\), where \(K_{\hh}(\x) = |\hh|^{-1/2}K(\hh^{-1/2}\x)\), \(K\) is a non-negative and radially symmetric kernel function that satisfies \(\int K(\x)\,\textrm{d}\x=1\), and \(\hh\) is a diagonal \(3\times3\) matrix of bandwidths that serve as smoothing parameters. The estimate \(\hf_{\Z}(\z) = \psi(\z, \hR(\z))\) aims at minimizing a locally weighted Kullback-Leibler distance to \(f_{\Z}(\z)\), and we refer to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) and Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for much more details about this construction, and to Section \ref{chap:asymptotic} for the relevant asymptotic estimation theory.

@@

We transform the observations to the \(z\)-scale as in \eqref{eq:pseudoobservations}, but now estimate the individual components in \(\R(\z)\) one by one, taking only into consideration the \emph{pair} of variables in question:
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
\label{eq:locallik2}
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices \((j,k)\), \(\psi(\cdot, \rho)\) is the bivariate version of \eqref{eq:lgdeapprox}, and \(\hh\) now is a \(2\times2\) diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \eqref{eq:localpartialcov} and \eqref{eq:estimate} as above.

@@

Equations \eqref{eq:matrixlocalpartial}, \eqref{eq:definition} and their empirical counterparts \eqref{eq:localpartialcov} and \eqref{eq:estimate} demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}). Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
\label{eq:popfixedh3}
\end{equation}
where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\). Denote by \(\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}\) the vector of local correlations. The joint limiting distribution of \(\hfrho\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):

\begin{theorem}
\protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\). Assume that the following conditions hold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  For any sequence of bandwidth matrices \(\hh_n\) tending to zero element-wise, there exists for the trivariate marginally standard Gaussian vector \(\Z\) a unique set of local correlations \(\frho_{\hh}(\z)\) that satisfies \eqref{eq:popfixedh3}, and there exists a \(\frho_0(\z)\) such that \(\frho_{\hh}(\z) \rightarrow \frho_0(\z)\).
\item
  The process generating \(\{\Z_t\}_{t=1}^n\) is \(\alpha\)-mixing with mixing coefficients \(\alpha(m)\) satisfying \(\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty\) for some \(\lambda > 1-2/\delta\) and \(\delta>2\).
\item
  \(n\rightarrow\infty\), and each bandwidth \(b = b_n\) tends to zero such that \(nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})\) for some constant \(\epsilon_0 > 0\).
\item
  In a given point \(\z\), the parameter space \(\Theta\) for each local correlation \(\rho(\z)\) is a compact subset of \((-1,1)\).
\item
  The non-negative kernel function \(K(\cdot)\) sasisfies \(\sup_{\z}K(\z) < \infty\), \(\int K(\y)\,\textrm{d}\y < \infty\), \(\partial/\partial z_iK(\z) <\infty\), and \(\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0\) for \(i = 1,2,3\).
\end{enumerate}

Then, by writing \(b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0\) assuming they all converge to zero at the same rate,
\begin{equation}
\sqrt{nb^5}\Jb\Mb^{-1/2}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
\label{eq:an3}
\end{equation}
where \(\bm{I}\) is the \(3\times3\) identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
\end{theorem}

Theorem \ref{thm:loccor3} and its assumptions above are variations of the assumptions and the results presented by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), and in particular Theorem 3 by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}). The analysis of the asymptotic behavior of \(\Jb\) and \(\Mb\) in the latter reference reveals that the local parameter estimates converge more slowly than, for example, the trivariate kernel density estimator (that converges as \(1/(nb^3)^{1/2}\)), and it is a straightforward exercise to modify their proof to suit our particular situation, which demonstrates that the three local correlations converge jointly in distribution at the rate \(1/(nb^5)^{1/2}\), see Section B in the supplement for details.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The locally Gaussian partial correlation},
  pdfauthor={Håkon Otneim and Dag Tjøstheim},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}

@@

\definecolor{Gray}{gray}{0.9}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

@@

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
\begin{document}
\maketitle
\begin{abstract}
It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally, we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
\end{abstract}

@@

Let \(\X = (X_1, \ldots, X_p)\) be a random vector. Denote by \((\X_1,\X_2,\X_3)\) a partition of \(\X\) into vectors of dimension \(d_1\), \(d_2\) and \(d_3\) respectively, such that \(\Xone = (\X_1, \X_2) = (X_1, \ldots,X_{d_1 + d_2})\) consists of the first \(d_1+d_2\) components in \(\X\) and \(\Xtwo = \X_3 = (X_{d_1+d_2+1}, \ldots, X_{d_1+d_2+d_3})\) contains the remaining \(d_3\) variables, where \(p = d_1+d_2+d_3\). We assume that the mean vector \(\fmu\) and covariance matrix \(\fSigma\) of \(\X\) exist, and partition them correspondingly, writing
\begin{equation}
\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix} \,\,\,\, \textrm{ and } \,\,\,\, \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33}\end{pmatrix},
\label{eq:matrixpartition}
\end{equation}
where \(\fSigma_{ij}\) is the covariance matrices of \((\X_i, \X_j)\), \(i,j = 1,2,3\). There are two main concepts of correlation when \(\Xtwo = \X_3\) is given, the \emph{partial} and the \emph{conditional} correlation, and they coincide in several joint distributions, among them the Gaussian. See e.g.~Baba, Shibata, and Sibuya (\protect\hyperlink{ref-baba2004partial}{2004}) for details. We will use the partial correlation as a starting point when defining the LGPC. The partial variance-covariance matrix of \(\Xone = (\X_1, \X_2)\) given \(\Xtwo = \X_3\) is
\begin{equation}
\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21},
\label{eq:matrixpartial}
\end{equation}
where
\[\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22} \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix}, \,\,\fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix} \, \textrm{ and } \, \fSigma^{22} = \fSigma_{33},\]
and \(\fSigma_{12|3}\) is the covariance matrix in the conditional (Gaussian) distribution of \(\Xone\) given \(\X_3\) if \(\X\) is jointly normal. The partial correlation matrix between \(\X_1\) and \(\X_2\) given \(\X_3\) is naturally defined as

@@

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
\label{eq:trans}
\end{equation}
The transformation enables us to simplify the local Gaussian approximation \eqref{eq:localgaussian} by writing the density \(f_{\Z}\) of \(\Z\) at the point \(\fv = \z\) as
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
\label{eq:lgdeapprox}
\end{equation}\\
where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a further simplified representation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.

@@

\[f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
where

@@

\begin{equation}
\hR(\z) = \arg\max_{\R} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \R) - \int K_{\hh}(\y - \z)\psi(\y, \R)\,\textrm{d}\y
\label{eq:locallik3}
\end{equation}
in each point \(\z\), where \(K_{\hh}(\x) = |\hh|^{-1}K(\hh^{-1}\x)\), \(K\) is a non-negative and radially symmetric kernel function that satisfies \(\int K(\x)\,\textrm{d}\x=1\), and \(\hh\) is a diagonal \(3\times3\) matrix of bandwidths that serve as smoothing parameters. The estimate \(\hf_{\Z}(\z) = \psi(\z, \hR(\z))\) aims at minimizing a locally weighted Kullback-Leibler distance to \(f_{\Z}(\z)\), and we refer to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) and Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for much more details about this construction, and to Section \ref{chap:asymptotic} for the relevant asymptotic estimation theory.

@@

We transform the observations to the \(z\)-scale as in \eqref{eq:pseudoobservations}, but now estimate the individual components in \(\R(\z)\) one by one, taking only into consideration the \emph{pair} of variables in question:
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
\label{eq:locallik2}
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices \((j,k)\), \(\psi(\cdot, \rho)\) is the bivariate version of \eqref{eq:lgdeapprox}, and \(\hh\) now is a \(2\times2\) diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \eqref{eq:localpartialcov} and \eqref{eq:estimate} as above.

@@

Equations \eqref{eq:matrixlocalpartial}, \eqref{eq:definition} and their empirical counterparts \eqref{eq:localpartialcov} and \eqref{eq:estimate} demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}). Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
\label{eq:popfixedh3}
\end{equation}
where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\) (at the point \(\z\)), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\). Denote by \(\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}\) the vector of local correlations. The joint limiting distribution of \(\hfrho\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):

\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\). Assume that the following conditions hold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  For any sequence of bandwidth matrices \(\hh_n\) tending to zero element-wise, there exists for the trivariate marginally standard Gaussian vector \(\Z\) a unique set of local correlations \(\frho_{\hh}(\z)\) that satisfies \eqref{eq:popfixedh3}, and there exists a \(\frho_0(\z)\) such that \(\frho_{\hh}(\z) \rightarrow \frho_0(\z)\).
\item
  The process generating \(\{\Z_t\}_{t=1}^n\) is \(\alpha\)-mixing with mixing coefficients \(\alpha(m)\) satisfying \(\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty\) for some \(\lambda > 1-2/\delta\) and \(\delta>2\).
\item
  \(n\rightarrow\infty\), and each bandwidth \(b = b_n\) tends to zero such that \(nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})\) for some constant \(\epsilon_0 > 0\).
\item
  In a given point \(\z\), the parameter space \(\Theta\) for each local correlation \(\rho(\z)\) is a compact subset of \((-1,1)\).
\item
  The non-negative kernel function \(K(\cdot)\) satisfies \(\sup_{\z}K(\z) < \infty\), \(\int K(\y)\,\textrm{d}\y < \infty\), \(\partial/\partial z_iK(\z) <\infty\), and \(\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0\) for \(i = 1,2,3\).
\end{enumerate}

Then, by writing \(b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0\) assuming they all converge to zero at the same rate,
\begin{equation}
\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
\label{eq:an3}
\end{equation}
where \(\bm{I}\) is the \(3\times3\) identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
\EndKnitrBlock{theorem}

It is shown in Section B of the Supplement that the matrices \(\Jb\) and \(\Mb\) are both of order \(b^2\). Theorem \ref{thm:loccor3} and its assumptions above are variations of the assumptions and the results presented by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), and in particular Theorem 3 by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}). The analysis of the asymptotic behavior of \(\Jb\) and \(\Mb\) in the latter reference reveals that the local parameter estimates converge more slowly than, for example, the trivariate kernel density estimator (that converges as \(1/(nb^3)^{1/2}\)), and it is a straightforward exercise to modify their proof to suit our particular situation, which demonstrates that the three local correlations converge jointly in distribution at the rate \(1/(nb^5)^{1/2}\), see Section B in the supplement for details.
